Factors affecting reliablity of state-space age-structured assessment models

### 2025-04-08 Project_0/manuscript/manuscript.tex
internal review for project_0 paper

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{12 March, 2025}

@@

\(^1\)corresponding author:
\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
\(^2\)Northeast Fisheries Science Center, Woods Hole Laboratory, 166
Water Street, Woods Hole, MA 02543 USA\\
\(^3\)Woods Hole Oceanographic Institution\\
\(^4\)SMAST\\
\(^5\)Institute of Marine Research\\
\(^6\)Mid-Atlantic Fisheries Management Council\\
\(^7\)Rutgers University\\

@@

@@

State-space assessment models, which are increasingly being used for
management of commercially important fish stocks, can include process
errors in many ways, and an evaluation of the statistical reliability of
the estimates of management output and best practices of selection among
alternative process error configurations are needed. We simulated 72
operating models with varying fishing pressure history, observation
error error, and process error magnitude, correlation, and source
(recruitment, survival, fishery selectivity, catchability, and natural
mortality). We fit estimating models with different assumptions on the
process error source, whether median natural mortality was estimated,
and whether a stock-recruit relationship was estimated. Estimating
models that assumed the correct process error source and median natural
mortality rate was known, and did not estimate a stock recruit
relationship generally had high probability of convergence and low bias
in terminal year SSB estimation. Bias was also low under many incorrect
process error assumptions when there was contrast in fishing pressure
and low observation error. Stock-recruitment parameters were not
reliably estimated in any scenarios, but the least accurate estimates
occured under ideal situations of low observation error, low recruitment
variability, and temporal contrast in fishing pressure. However,
estimation of median natural mortality rate was reliable for many OMs
(other than those with survival process errors) with temporal contrast
in fishing pressure regardless of the estimated process error type.
Marginal AIC most accurately distinguished process errors on
recruitment, survival, and selectivity, as well as larger magnitude
process errors of other types. Retrospective patterns for spawning stock
biomass and fishing mortality were generally weak, but those for
recruitment were sizable when observation error was high, even with the
correct process error assumptions.

@@

Application of state-space models in fisheries stock assessment and
management has expanded dramatically within International Council for
the Exploration of the Sea (ICES), Canada, and the Northeast US (Nielsen
and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and Miller
2021). State-space models latent population characteristics as
statistical time series with periodic observations that also may have
error due to sampling or other sources of measurement error. Traditional
assessment models may use state-space approaches to account for temporal
variability in population characteristics (Legault and Restrepo 1999;
Methot and Wetzel 2013), but these models treat the annual parameters as
penalized fixed effects parameters where the variance parameters
controlling the penalties are assumed known (Thorson and Minto 2015).
Modern state-space models can estimate the annually varying parameters
as random effects with variance parameters estimated using maximum
marginal likelihood or corresponding Bayesian approaches. These latter
approaches are considered best practice and a requirement for the next
generation of stock assessment models (Hoyle et al. 2022; Punt 2023).

@@

Despite mounting efforts, several limitations remain. First, confounding
processes that can be treated as random effects in the model were not
thoroughly examined or tested within a simulation-estimation framework.
Second, previous studies relied on operating models conditioned on
specific fisheries, limiting their generalizability (Li et al. In
review; Liljestrand et al. 2024). In particular, the effects of
observation error and underlying fishing history were not fully isolated
from their study design, making it challenging to disentangle the
interplay between process and observation error magnitudes, as
demonstrated in Fisch et al. (2023). Third, explicitly modeling
stock-recruitment relationships as mechanistic drivers of population
dynamics is promising (Fleischman et al. 2013; Du Pontavice et al. 2022)
but can be challenging in state-space models. Evidence suggests that
when both process and observation errors are unknown, estimating density
dependence parameters becomes highly uncertain (Knape 2008; Polansky et
al. 2009). In particular, Knape (2008) demonstrated that stronger
density dependence becomes increasingly difficult to estimate in the
presence of observation error. Therefore, it is crucial to assess
whether density dependence mechanisms can be estimated with sufficient
precision before drawing ecological conclusions (Auger-Méthé et al.
2016).

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a stock-recruit model
was estimated, and whether a constant, or, in some EMs, median, natural
mortality is estimated. Given the confounding nature of process errors,
developing diagnostic tools to detect model misspecification is of great
scientific interest and could aid the next generation of stock
assessments (Auger-Méthé et al. 2021). We evaluate whether convergence
and Akaike Information Criterion (AIC) can correctly determine the
source of process error and the existence of a stock-recruit
relationship. We also evaluate when retrospective patterns occur and the
degree of bias in the outputs of the assessment model that are important
for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
in our simulation study (Miller and Stock 2020; Stock and Miller 2021).
WHAM is an R package freely available as a github repository. For this
study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. This package has also been used to
configure OMs and EMs for closed loop simulations evaluating index-based
assessment methods (Legault et al. 2023) and is used for management of
haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea
bass, and yellowtail flounder in the Northeast US.

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects are assumed:
abundance at age (R, R+S), natural mortality (R+M), fleet selectivity
(R+Sel), or index catchability (R+q). For each OM assumption about
variance of process errors and observations are required and the values
we used were based on a review of the range of estimates from
applications of WHAM in management of stocks of haddock, butterfish, and
American plaice in the NE US.

In total, we configured 72 OMs with alternative assumptions about the
source and variability of process errors, level of observation error in
indices and age composition data, and contrast in fishing pressure over
time. We fitted 20 EMs to observations from each of 100 simulations
where process errors were also simulated. For R+M, R+Sel, and R+q OMs,
we used unique seeds for each simulation, but we inadvertently used the
same 100 seeds for all R and R+S OMs. Each of the EMs made alternative
assumptions about the source of process errors and whether natural
mortality (or the median for models with process error in natural
mortality) was estimated and whether a Beverton-Holt stock recruit
relationship was estimated within the EM. Details of each of the
operating and EMs are described below.

We did not use the log-normal bias-correction feature for process errors
or observations described by (Stock and Miller 2021) for operating and
EMs (Li et al. In review). Simulations and model fitting were all
carried out on the University of Massachusetts Green High-Performance
Computing Cluster. All code we used to perform the simulation study and
summarize results can be found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

The population consists of 10 age classes: ages 1 to 10+ and we assume
spawning occurs each year 1/4 of the way through the year. The maturity
at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
(Figure \ref{om_inputs_fig}, top left).

@@

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all OMs. All post-recruit
productivity components are constant in the NAA and survey catchability
process error OMs. Therefore steepness and unfished recruitment are also
constant over the time period for those OMs (Miller and Brooks 2021). We
specified unfished recruitment = \(R_0 = e^{10}\) and
\(\Fmsy = F_{40\%} = 0.348\) equated to a steepness of 0.69 and
\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the \[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\] Beverton-Holt parameterization (Figure \ref{om_inputs_fig}, bottom
right). For OMs without process errors on natural mortality we assumed
the rate was assumed 0.2. For OMs with process errors on natural
mortality the median natural mortality rate was 0.2.

We used two fishing scenarios for OMs. In the first scenario, the stock
experiences overfishing at 2.5\Fmsy for the first 20 years and fishing
at \Fmsy for the last 20 years (denoted \(2.5\Fmsy \rightarrow \Fmsy\)).
In the second scenario, the stock is fished at \Fmsy for the entire time
period. The magnitude of the overfishing assumptions is based on average
estimates of overfishing for NE groundfish stocks from (Wiedenmann et
al. 2019). Legault et al. (2023) also used similar approaches to
defining fishing mortality histories for OMs.

We specified initial population abundance at age at the equilibrium
distribution fishing at either \(F = 2.5\times \Fmsy\) or \(F = \Fmsy\)
for the two alternative fishing histories. This implies that, for a
deterministic model, the abundance at age would not change from the
first year to the next.

For OMs with time-varying random effects for M, steepness is not
constant, but we used the same alpha and beta parameters as other OMs
this equates to a steepness and R0 at the median of the time series
process for M. For OMs with time-varying random effects for fishery
selectivity, \Fmsy is also not constant however we use the same F
history as other OMs which corresponds to Fmsy at the mean selectivity
parameters.

@@

We assumed a single fleet operating year round for catch observations
with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
1 (Figure \ref{om_inputs_fig}, bottom left). This selectivity is was
used to define \Fmsy for the Beverton-Holt stock recruitment parameters
above. We assumed a logistic-normal distribution with no correlation on
the multivariate normal scale for the age-composition observations for
the fleet.

@@

Two time series of surveys are assumed and observed in numbers rather
than biomass for the entire 40 year period with one occurring in the
spring (0.25 of each year) and one in the fall (0.75 of each year).
Catchability of both surveys are assumed to be 0.1. Like the fishing
fleet, we assumed logistic selectivity for both indices with
\(a_{50} = 5\) and slope = 1 and a logistic-normal distribution with no
correlation on the multivariate normal scale for the age-composition
observations.

@@

For operating models with random effects on recruitment and(or) survival
(R, R+S) we assumed marginal standard deviations for recruitment of
\(\sigma_R \in \{0.5,1.5\}\) and marginal standard deviations for older
age classes of \(\sigma_{2+} \in \{0,0.25, 0.5\}\). The full factorial
combination of these process error assumptions and fishing history (2
levels) and observation error (2 levels) scenarios described above
results in 24 different R (\(\sigma_{2+} = 0\)) and R+S operating models
(Table \ref{naa_om_table}).

@@

All R+M OMs treat natural mortality constant across age, but with
annually varying random effects. WHAM treats natural mortality as a
log-transformed parameter \[
\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
\] that is a linear combination of a mean log-natural mortality
parameter that is constant across ages \(\mu_{M} = \log(0.2)\) and any
annual random effects marginally distributed as
\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). Uncorrelated
random effects were also included on recruitment with \(\sigma_R = 0.5\)
(hence, R+M). The marginal standard deviations we assumed for log
natural mortality random effects were \(\sigma_M \in \{0.1, 0.5\}\) and
AR1 autocorrelation parameters of \(\rho_M \in \{0,0.9\}\). The full
factorial combination of these process error assumptions and fishing
history (2 levels) and observation error (2 levels) scenarios described
above results in 16 different R+M OMs (Table \ref{M_om_table}).

@@

MORE SPECIFICS about correlation of random effects? Both selectivity
pars? just correlated by year? WHAM treats selectivity parameter \(s\)
as a logit-transformed parameter \[
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
\] that is a linear combination of a mean \(\mu_s\) and any annual
random effects marginally distributed as
\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\) where the
lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
specified by the user. All selectivity parameters are either \(a_50\) or
slope parameters and we assume bounds of 0 and 10 for all selectivity
parameters for all operating and EMs. The marginal standard deviations
we assumed for logit scale random effects were
\(\sigma_s \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
\(\rho_s \in \{0,0.9\}\). Like R+M OMs, the full factorial combination
of these process error assumptions and fishing history (2 levels) and
observation error (2 levels) scenarios described above results in 16
different R+Sel OMs (Table \ref{sel_om_table}).

@@

Like selectivity parameters, WHAM treats catchability for an index \(i\)
as a logit-transformed parameter \[
\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
\] that is a linear combination of a mean \(\mu_i\) and any annual
random effects marginally distributed as
\(\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)\) where the
lower and upper bounds of the catchability (\(l_i\) and \(u_i\)) can be
specified by the user. Here we assume bounds of 0 and 1000 for all
operating and EMs. For operating and EMs with process errors on
catchability, the temporal variation is only assumed for the first
index. The marginal standard deviations we assumed for logit scale
random effects were \(\sigma_i \in \{0.1, 0.5\}\) and AR1
autocorrelation parameters of \(\rho_i \in \{0,0.9\}\). Like R+M and
R+Sel OMs, the full factorial combination of these process error
assumptions and fishing history (2 levels) and observation error (2
levels) scenarios described above results in 16 different R+q OMs (Table
\ref{q_om_table}).

@@

For each data set simulated from an OM 20 EMs were fit. A total of 32
different EMs were fit across OMs where the subset of 20 depended on the
source of process error in the OM (Table \ref{em_table}). The EMs make
different assumptions about the source of process error (R+S, R+M,
R+Sel, R+q) and whether there is temporal autocorrelation or not, and
whether a Beverton-Holt stock recruit function is estimated and the
natural mortality rate (\(\mu_M\), the constant or mean on log scale for
R+M EMs) is estimated. For simplicity we refer to the derived estimate
\(e^{\mu_M}\) as the median natural mortality rate regardless of whether
natural mortality random effects are estimated in the EM. The first 20
EMs in Table \ref{em_table} were fit to simulate data sets from R and
R+S OMs. EMs 5 to 24 in Table \ref{em_table} were fit to simulate data
sets from R+M OMs. EMs 5 to 20 and 25-28 in Table \ref{em_table} were
fit to simulate data sets from R+Sel OMs. Finally, EMs 5 to 20 and 29-32
in Table \ref{em_table} were fit to simulate data sets from R+q OMs. The
maturity at age, weight at age for catch and SSB, and observation error
variance of aggregate catch and indices were all assumed known at the
true values.

@@

The first measure of reliability we investigated was frequency of
convergence when fitting each EM to the simulated data sets. There are
various ways to assess convergence of the fit (e.g., Carvalho et al.
2021), but given the importance of estimates of uncertainty when using
assessment models in management, we estimated probablity of convergence
as measured by occurrence of a positive-definite hessian matrix at the
optimized negative log-likelihood that could be inverted. However, we
also provided results in the Supplementary Materials for the maximum of
the absolute values among all gradients for all fits of a given EM to
all simulated data sets from a given OM that produced hessian-based
standard errors for all estimated fixed effects. This provides an
indication of how poor the calculated gradients can be, but still
presumably converged adequately enough for parameter inferences.

@@

We estimated the probability of selection of each process error model
structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given
operating model, we compared AIC for EMs that all made the same
assumptions about median natural mortality (known or estimated) and
stock recruitment model (Beverton-Holt or none).

We also estimated the probability of correctly selecting between models
with Beverton-Holt stock recruit function assumed and models without the
stock-recruit function (null model). We made these comparisons between
models that otherwise assume the same process error structure as the
operating model and both of the compared models either estimate median
natural mortality or assume it is known. Contrast in fishing pressure
and time series with recruitment at low stock size has been shown to
improve estimability of stock-recruitment parameters (Magnusson and
Hilborn 2007; Conn et al. 2010). Our preliminary inspections of the
proportions of simulations where the correct recruitment model was
chosen for a given set of OM factors (including contrast in fishing
pressure) indicated generally poor performance of AIC. Therefore, we fit
logistic regression models to the indicator of Beverton-Holt models
having lower AIC as a function of the log-standard deviation of the true
log(SSB) (similar to the log of the coefficient of variation for SSB)
since simulations with realized SSB producing low and high recruitments
would have larger variation in realized SSB.

All results only condition on whether all of the compared estimating
models completed the optimization process without failure. We did not
condition on convergence as defined by a gradient threshold or
invertibity of the hessian because optimization can correctly determine
the the correct likelihood that would indicate poor convergence because
variance parameters may be at the lower bound of zero correctly for
models that assume the incorrect process error structure.

@@

\hypertarget{convergence}{%
\subsection*{Convergence}\label{convergence}}
\addcontentsline{toc}{subsection}{Convergence}

>>>>

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{08 April, 2025}

@@

\(^1\)corresponding author:
\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
\(^2\)Northeast Fisheries Science Center, Woods Hole Laboratory, 166
Water Street, Woods Hole, MA 02543 USA\\
\(^3\)Woods Hole Oceanographic Institution\\
\(^4\)SMAST\\
\(^5\)Mid-Atlantic Fishery Management Council, 800 North State Street,
Suite 201, Dover, DE 19901 USA\\
\(^6\)Institute of Marine Research, Nye Flødevigveien 20, 4817 His,
Norway\\
\(^7\)Department of Ecology, Evolution, and Natural Resources. Rutgers
University\\

@@

\textbf{keywords:} stock assessment, state-space, model selection, bias,
convergence, retrospective patterns

@@

State-space models are increasingly used for stock assessment, and
evaluations of their statistical reliability and best practices for
selecting among process error configurations are needed. We simulated 72
operating models that varied fishing pressure and observation error
across process errors in recruitment, survival, selectivity,
catchability, and/or natural mortality. We fit estimating models with
different assumptions on the process error source and whether median
natural mortality or a stock-recruit relationship were estimated.
Estimating models without a stock-recruit relationship that assumed the
correct process error source and median natural mortality had high
convergence rates and low bias. Bias was also low under many incorrect
process error assumptions when there was contrast in fishing pressure
and low observation error. Marginal AIC most accurately distinguished
process errors on recruitment, survival, and selectivity, as well as
larger magnitude process errors of other types. Retrospective patterns
were generally small but were sizable for recruitment when observation
error was high. These results help establish the statistical reliability
of state space assessment models and pave the way for the
next-generation of fisheries stock assessment

@@

Application of state-space models in fisheries stock assessment and
management has expanded dramatically within International Council for
the Exploration of the Sea (ICES), Canada, and the Northeast US (Nielsen
and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and Miller
2021). State-space models latent population characteristics as
statistical time series with periodic observations that also may have
error due to sampling or other sources of measurement error. Traditional
assessment models may use state-space approaches to account for temporal
variability in population characteristics (Legault and Restrepo 1999;
Methot and Wetzel 2013), but these models treat the annual parameters as
penalized fixed effects parameters where the variance parameters
controlling the penalties are assumed known (Thorson and Minto 2015).
Modern state-space models can estimate the annually varying parameters
as random effects with variance parameters estimated using maximum
marginal likelihood or corresponding Bayesian approaches. These latter
approaches are considered best practice and a recommended for the next
generation of stock assessment models (Hoyle et al. 2022; Punt 2023).

@@

Despite mounting efforts, several limitations remain. First, confounding
processes that can be treated as random effects in the model were not
thoroughly examined or tested within a simulation-estimation framework.
Second, previous studies relied on operating models conditioned on
specific fisheries, limiting their generalizability (Li et al. In
review; Liljestrand et al. 2024). In particular, the effects of
observation error and underlying fishing history have not been fully
isolated in simulation study designs, making it challenging to
disentangle the interplay between process and observation error
magnitudes, as demonstrated in Fisch et al. (2023). Third, explicitly
modeling stock-recruit relationships (SRRs) as mechanistic drivers of
population dynamics is promising (Fleischman et al. 2013; Du Pontavice
et al. 2022), but can reliability of inferences within integrated
state-space age-structured models has not been evaluated. Evidence from
other studies suggests that when both process and observation errors are
unknown, estimating density dependence parameters becomes highly
uncertain (Knape 2008; Polansky et al. 2009). In particular, Knape
(2008) demonstrated that stronger density dependence becomes
increasingly difficult to estimate in the presence of observation error.
Therefore, it is crucial to assess whether density dependence mechanisms
can be estimated with sufficient precision for use in fisheries
management (Auger-Méthé et al. 2016).

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a SRR was estimated,
and whether a constant, or, in some EMs, median, natural mortality is
estimated. Given the confounding nature of process errors, developing
diagnostic tools to detect model misspecification is of great scientific
interest and could aid the next generation of stock assessments
(Auger-Méthé et al. 2021). We evaluate whether convergence and Akaike
Information Criterion (AIC) can correctly determine the source of
process error and the existence of a SRR. We also evaluate when
retrospective patterns occur and the degree of bias in the outputs of
the assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
in our simulation study (Miller and Stock 2020; Stock and Miller 2021).
WHAM is an R package freely available via a github repository and is
built on the Template Model Builder package (Kristensen et al. 2016).
For this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. WHAM has also been used to configure OMs
and EMs for closed loop simulations evaluating index-based assessment
methods (Legault et al. 2023) and is currently used or accepted for use
in management of numerous fish stocks in the Northeast United States
(NEUS) (e.g., NEFSC 2022a, 2022b; NEFSC 2024).

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects were assumed:
recruitment (R, assumed present in all models), apparent survival
(denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
index catchability (R+q). We refer to the (R+S) OMs as modeling apparent
survival because on logscale the random effects (\(\epsilon_{a,y}\)) are
additive to the total mortality (F+M) between numbers at age, thus they
modify the survival term. However, as Stock and Miller (2021) note,
these random effects can be due to events other than mortality, such as
immigration, emigration, missreported catch, and other sources of
misspecification. For each OM, assumptions about the magnitude of the
variance of process errors and observations are required and the values
we used were based on a review of the range of estimates from NEUS
assessments using WHAM in stocks NEUS.

In total, we configured 72 OMs with alternative assumptions about the
source and magnitude of process errors, magnitude of observation error
in indices and age composition data, and contrast in fishing pressure
over time. We fitted 20 EMs to observations generated from each of 100
simulations where process errors were also simulated. Each EM differed
in assumptions about the source of process errors, whether natural
mortality (or the median for models with process error in natural
mortality) was estimated, and whether a Beverton-Holt SRR was estimated
within the EM. Details of each of the OMs and EMs are described below.

We did not use the log-normal bias-correction feature for process errors
or observations described by (Stock and Miller 2021) for OMs and EMs to
simplify interpretation of the study results (Li et al. In review).
Simulations and model fitting were all carried out on the University of
Massachusetts Green High-Performance Computing Cluster. All code we used
to perform the simulation study and summarize results can be found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

The population consists of 10 age classes, ages 1 to 10+, with the last
being a plus group that accumulates ages 10 and older. We assume
spawning occurs annually 1/4 of the way through the year. The maturity
at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
(Figure \ref{om_inputs_fig}, top left).

@@

We assumed a Beverton-Holt SRR with constant pre-recruit mortality
parameters for all OMs. All post-recruit productivity components are
constant in the apparent survival (R+S) and survey catchability (R+Sel)
process error OMs. Therefore, steepness and unfished recruitment are
also constant over the time period for those OMs (Miller and Brooks
2021). We specified unfished recruitment equal to \(e^{10}\) and
\(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69 and
\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
parameterization \[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\] (Figure \ref{om_inputs_fig}, bottom right). For OMs without process
errors on natural mortality we fixed the rate at 0.2. For OMs with
process errors on natural mortality, the median natural mortality rate
was specified to be 0.2.

We used two fishing scenarios for OMs. In the first scenario, the stock
experiences overfishing at 2.5\Fmsy for the first 20 years followed by
fishing at \Fmsy for the last 20 years (denoted
\(2.5\Fmsy \rightarrow \Fmsy\)). In the second scenario, the stock is
fished at \Fmsy for the entire time period (40 years). The magnitude of
the overfishing assumptions is based on average estimates of overfishing
for NEUS groundfish stocks from Wiedenmann et al. (2019) and similar to
the approach in Legault et al. (2023).

We specified initial population abundance at age at the equilibrium
distribution that corresponds to fishing at either
\(F = 2.5\times \Fmsy\) or \(F = \Fmsy\). This implies that, for a
deterministic model, the abundance at age would not change from the
first year to the next.

For OMs with time-varying random effects for M , steepness is not
constant. However, we used the same \(a\) and \(b\) parameters as other
OMs, which equates to a steepness and R0 at the median of the time
series process for M. For OMs with time-varying random effects for
fishery selectivity, \Fmsy is also not constant, but since we use the
same F history as other OMs, this corresponds to \Fmsy at the mean
selectivity parameters.

@@

We assumed a single fleet operating year round for catch observations
with logistic selectivity for the fleet (\(a_{50} = 5\) and slope = 1;
Figure \ref{om_inputs_fig}, bottom left). This selectivity was used to
define \Fmsy for the Beverton-Holt SRR parameters above. We assumed a
logistic-normal distribution with no correlation on the multivariate
normal scale for the age-composition observations for the fleet.

@@

Two time series of fishery-independent surveys in numbers are generated
for the entire 40 year period with one occurring in the spring (0.25 of
each year) and one in the fall (0.75 of each year). Catchability of both
surveys are assumed to be 0.1. Like the fishing fleet, we assumed
logistic selectivity for both indices (\(a_{50} = 5\) and slope = 1) and
a logistic-normal distribution with no correlation on the multivariate
normal scale for the age-composition observations.

@@

For operating models with random effects on recruitment and(or) apparent
survival (R, R+S), we assumed marginal standard deviations for
recruitment of \(\sigma_R \in \{0.5,1.5\}\) and marginal standard
deviations for older age classes of \(\sigma_{2+} \in \{0,0.25, 0.5\}\).
The full factorial combination of these process error assumptions (2x3
levels) and scenarios for fishing history (2 levels) and observation
error (2 levels) scenarios described above results in 24 different R
(\(\sigma_{2+} = 0\)) and R+S operating models (Table
\ref{naa_om_table}).

@@

All R+M OMs treat natural mortality as constant across age, but with
annually varying random effects. WHAM treats natural mortality as a
log-transformed parameter \[
\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
\] that is a linear combination of a mean log-natural mortality
parameter that is constant across ages (\(\mu_{M} = \log(0.2)\)) and any
annual random effects are marginally distributed as
\(\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)\). Uncorrelated
random effects were also included on recruitment with \(\sigma_R = 0.5\)
(hence, we denote these OMs as R+M). The marginal standard deviations we
assumed for log natural mortality random effects were
\(\sigma_M \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
\(\rho_M \in \{0,0.9\}\). The full factorial combination of these
process error assumptions and fishing history (2 levels) and observation
error (2 levels) scenarios described above results in 16 different R+M
OMs (Table \ref{M_om_table}).

@@

WHAM treats selectivity parameter \(s\) as a logit-transformed parameter
\[
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
\] that is a linear combination of a mean \(\mu_s\) and any annual
random effects marginally distributed as
\(\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)\), where the
lower and upper bounds of the parameter (\(l_s\) and \(u_s\)) can be
specified by the user. All selectivity parameters (\(a_{50}\) and slope
parameters) were bounded by 0 and 10 for all OMs and EMs. The marginal
standard deviations we assumed for logit scale random effects were
\(\sigma_s \in \{0.1, 0.5\}\) and AR1 autocorrelation parameters of
\(\rho_s \in \{0,0.9\}\). Like R+M OMs, the full factorial combination
of these process error assumptions (2x2 levels) and scenarios described
above for fishing history (2 levels) and observation error (2 levels)
results in 16 different R+Sel OMs (Table \ref{sel_om_table}).

@@

Like selectivity parameters, WHAM treats catchability for an index \(i\)
as a logit-transformed parameter \[
\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
\] that is a linear combination of a mean \(\mu_i\) and any annual
random effects marginally distributed as
\(\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)\) where the
lower and upper bounds of the catchability (\(l_i\) and \(u_i\)) can be
specified by the user. We assumed bounds of 0 and 1000 for all OMs and
EMs. For all OMs and EMs with process errors on catchability, the
temporal variation only applies to the first index, which could be
interpreted as capturing some unmeasured seasonal process that affects
availability to the survey. The marginal standard deviations we assumed
for logit scale random effects were \(\sigma_i \in \{0.1, 0.5\}\) and
AR1 autocorrelation parameters of \(\rho_i \in \{0,0.9\}\). Like R+M and
R+Sel OMs, the full factorial combination of these process error
assumptions and fishing history (2 levels) and observation error (2
levels) scenarios described above results in 16 different R+q OMs (Table
\ref{q_om_table}).

@@

For each of the data sets simulated from an OM, 20 EMs were fit. A total
of 32 different EMs were fit across OMs where the subset of 20 depended
on the source of process error in the OM (Table \ref{em_table}). The EMs
have different assumptions about the source of process error (R+S, R+M,
R+Sel, R+q) and whether or not 1) there is temporal autocorrelation, 2)
a Beverton-Holt SRR is estimated, and 3) the natural mortality rate
(\(\mu_M\), the constant or mean on log scale for R+M EMs) is estimated.
For simplicity we refer to the derived estimate \(e^{\mu_M}\) as the
median natural mortality rate regardless of whether natural mortality
random effects are estimated in the EM.

Subsets of 20 EMs in Table \ref{em_table} were fit to simulate data sets
from each of the OM process error categories. For R and R+S OMs, fitted
EMs had matching process error assumptions as well as R+Sel, R+M, and
R+q assumptions without autocorrelation. Similarly, For other OM process
error categories, we fit EMs with matching process error assumptions as
well as other process error types without autocorrelation. The maturity
at age, weight at age for catch and SSB, and observation error variance
of aggregate catch and indices were all assumed known at the true
values. However, the variance parameters for the logistic-normal
distributions for age composition observations were estimated in the
EMs.

@@

\hypertarget{convergence}{%
\subsubsection{Convergence}\label{convergence}}

The first measure of reliability we investigated was frequency of
convergence when fitting each EM to the simulated data sets. There are
various ways to assess convergence of the fit (e.g., Carvalho et al.
2021; Kapur et al. 2025), but given the importance of estimates of
uncertainty when using assessment models in management, we estimated
probablity of convergence as measured by occurrence of a
positive-definite hessian matrix at the optimized negative
log-likelihood that could be inverted. We also provide results in the
Supplementary Materials for the maximum of the absolute values among all
gradients for all fits of a given EM to all simulated data sets from a
given OM that produced hessian-based standard errors for all estimated
fixed effects. This provides an indication of how poor the calculated
gradients can be, but still presumably converged adequately enough for
parameter inferences. We used the Clopper-Pearson exact method for
constructing 95\% confidence intervals of the probabilities of
convergence (Clopper and Pearson 1934; Thulin 2014).

@@

We estimated the probability of selection of each process error model
structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given
operating model, we compared AIC for EMs that all made the same
assumptions about median natural mortality (known or estimated) and SRR
(Beverton-Holt or none).

We also estimated the probability of correctly selecting between EMs
with Beverton-Holt SRR assumed and models without the SRR (null model).
We made these comparisons between models that otherwise assumed the same
process error structure as the operating model and both of the compared
models either estimate median natural mortality or assume it is known.
Contrast in fishing pressure and time series with recruitment at low
stock size has been shown to improve estimation of SRR parameters
(Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
inspections of the proportions of simulations where the correct
recruitment model was chosen for a given set of OM factors (including
contrast in fishing pressure) indicated generally poor performance of
AIC. Therefore, we fit logistic regression models to the indicator of
Beverton-Holt models having lower AIC as a function of the log-standard
deviation of the true log(SSB) (similar to the log of the coefficient of
variation for SSB) since simulations with realized SSB producing low and
high recruitments would have larger variation in realized SSB.

All model selection results condition on whether all of the compared
estimating models completed the optimization process without failure. We
did not condition on convergence as defined by a gradient threshold or
invertibity of the hessian because optimization could correctly
determine an inappropriate process error assumption by estimating
variance parameters at the lower bound of zero. Such an optimization
could indicate poor convergence but the likelihood would be equivalent
to that without the mis-specified random effects and the AIC would be
appropriately higher because more (variance) parameters were estimated.
All other measures of reliability described below (bias and Mohn's
\(\rho\)) use these same criteria for inclusion of EM fits in the
summarized results.

@@

\hypertarget{convergence-1}{%
\subsection*{Convergence}\label{convergence-1}}
\addcontentsline{toc}{subsection}{Convergence}
